A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-27 inst/manuscript/manuscript.Rmd
reanalyzing simulation data

source(here("R", "sim_pop.R"))
N <- 9
pop_df <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                  psill = 0.9, erange = 1, nugget = 0.1)

@@

pop_df_mod <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                      psill = 0.9, erange = 1, nugget = 0.1)
pop_df_mod2 <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                       psill = 0.9, erange = 1, nugget = 0.1)
pop_df_mod3 <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                      psill = 0.9, erange = 1, nugget = 0.1)

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four approaches from Table \ref{tab:designanalysis} with "IRS-Design" as the baseline:
\mbox{}
\begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

@@

Unsurprisingly, Figure \ref{fig:figeff} also shows that, when the assumptions for GRTS-Model are satisfied, the approach outperforms GRTS-Design. However, even when the model that generates the data is different than the model used to fit the data, as in the lognormal response, the model-based approach still outperforms the design-based approach when there is a high amount of spatial correlation.

@@

combo_data <- combo_data %>%
  dplyr::filter(n != 10 & gridded == FALSE) %>%
  mutate(sim = interaction(n, psill, resptype)) %>% ## , gridded
  mutate(approach = fct_relevel(approach, c("Design IRS", "Model IRS",
                                            "Design GRTS", "Model GRTS"))) %>%
  mutate(resptype = fct_relevel(resptype, c("normal", "lognormal")))
combo_data <- combo_data %>% group_by(sim) %>% mutate(designirsrmspe = if_else(approach == "Design IRS", true = rmspe, false = NA_real_)) %>%
  fill(designirsrmspe, .direction = "downup") %>%
  mutate(rel_efficiency = rmspe / designirsrmspe) %>%
  ungroup() %>%
  mutate(psill_ratio = 1 - nugget_ratio) %>%
  mutate(n_factor = factor(n))

@@

We also studied 95% interval coverage among the approaches. The design-based 95% confidence intervals and model-based 95% prediction intervals are constructed using the normal distribution. Justification for the design-based intervals lies in the asymptotic normality of totals via the Central Limit Theorem, and justification for the model-based intervals lies in the normality assumption of the errors.  Figure \ref{fig:figconf} shows the 95% interval coverage for each of the four approaches. All four approaches have somewhat similar interval coverage in all settings, with GRTS-Design having slightly lower coverage when the response is normal.

In the normal response settings, all approaches have coverage around 95%. This is expected, as the intervals are also based on the normal distribution In the lognormal response settings, however, all approaches have coverage below 95%. This is also expected, as the intervals are still based on the normal distribution. In the lognormal response settings, interval coverage increases both as the sample size increases and as the strength of spatial dependence increases. This suggests that the larger the sample size and the stronger the spatial dependence, the more resistant these intervals are to departures from normality of the data.

>>>>

source(here("R", "sim_pop.R"))
N <- 9
pop_df <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                  psill = 0.9, range = 1, nugget = 0.1)

@@

pop_df_mod <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                      psill = 0.9, range = 1, nugget = 0.1)
pop_df_mod2 <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                       psill = 0.9, range = 1, nugget = 0.1)
pop_df_mod3 <- sim_pop(N = N, gridded = TRUE, cortype = "Exponential",
                      psill = 0.9, range = 1, nugget = 0.1)

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four approaches from Table \ref{tab:designanalysis} using the random location layout, where "IRS-Design" is the baseline:
\mbox{}
\begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

@@

Unsurprisingly, Figure \ref{fig:figeff} also shows that, when the assumptions for GRTS-Model are satisfied, the approach outperforms GRTS-Design. However, even when the model that generates the data is different than the model used to fit the data, as in the lognormal response, the model-based approach still outperforms the design-based approach when there is a high amount of spatial correlation. Additionally, as the sample size increases, IRS-Design performs relatively worse compared to the other approaches. These conclusions were similar to those observed when the data were gridded.

@@

combo_data <- combo_data %>%
  dplyr::filter(gridded == FALSE) %>%
  mutate(sim = interaction(n, psill, resptype)) %>% ## , gridded
  mutate(approach = fct_relevel(approach, c("Design IRS", "Model IRS",
                                            "Design GRTS", "Model GRTS"))) %>%
  mutate(resptype = fct_relevel(resptype, c("normal", "lognormal")))
combo_data <- combo_data %>% group_by(sim) %>% mutate(designirsrmspe = if_else(approach == "Design IRS", true = rmspe, false = NA_real_)) %>%
  fill(designirsrmspe, .direction = "downup") %>%
  mutate(rel_efficiency = rmspe / designirsrmspe) %>%
  ungroup() %>%
  mutate(psill_ratio = 1 - nugget_ratio) %>%
  mutate(n_factor = factor(n))

@@

We also studied 95% interval coverage among the approaches. The design-based 95% confidence intervals and model-based 95% prediction intervals are constructed using the normal distribution. Justification for the design-based intervals lies in the asymptotic normality of totals via the Central Limit Theorem, and justification for the model-based intervals lies in the normality assumption of the errors.  Figure \ref{fig:figconf} shows the 95% interval coverage for each of the four approaches in the random location layout. All four approaches have somewhat similar interval coverage in all settings, with GRTS-Design having slightly lower coverage when the response is normal.

In the normal response settings, all approaches have coverage around 95%. This is expected, as the intervals are also based on the normal distribution In the lognormal response settings, however, all approaches have coverage below 95%. This is also expected, as the intervals are still based on the normal distribution. In the lognormal response settings, interval coverage increases both as the sample size increases and as the strength of spatial dependence increases. This suggests that the larger the sample size and the stronger the spatial dependence, the more resistant these intervals are to departures from normality of the data. These conclusions were similar to those observed when the data were gridded.

### 2021-10-27 inst/manuscript/manuscript.tex
reanalyzing simulation data

We used a crossed design with the simulation parameters given in Table
\ref{tab:parmtab} for a total of 36 scenarios. All scenarios used
exponential correlation with a \(\sqrt{2} / 3\) for \(N = 900\) response
values simulated on the unit square in either random locations (Layout =
Random) or gridded locations (Layout = Gridded). The range was chosen so
that the decayed to nearly zero at the largest distance possible in the
domain, \(\sqrt{2}\), otherwise known as the ``effective'' range (for
the exponential covariance, the effective range is \(3\phi\)). The mean
for the spatial process generating the response was set to zero.

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four
approaches from Table \ref{tab:designanalysis} with ``IRS-Design'' as
the baseline: \mbox{} \begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

@@

We also studied 95\% interval coverage among the approaches. The
design-based 95\% confidence intervals and model-based 95\% prediction
intervals are constructed using the normal distribution. Justification
for the design-based intervals lies in the asymptotic normality of
totals via the Central Limit Theorem, and justification for the
model-based intervals lies in the normality assumption of the errors.
Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four approaches. All four approaches have somewhat similar interval
coverage in all settings, with GRTS-Design having slightly lower
coverage when the response is normal.

In the normal response settings, all approaches have coverage around
95\%. This is expected, as the intervals are also based on the normal
distribution In the lognormal response settings, however, all approaches
have coverage below 95\%. This is also expected, as the intervals are
still based on the normal distribution. In the lognormal response
settings, interval coverage increases both as the sample size increases
and as the strength of spatial dependence increases. This suggests that
the larger the sample size and the stronger the spatial dependence, the
more resistant these intervals are to departures from normality of the
data.

>>>>

We used a crossed design with the simulation parameters given in Table
\ref{tab:parmtab} for a total of 36 scenarios. All scenarios used
exponential correlation with a \(\sqrt{4} / 3\) for \(N = 900\) response
values simulated on the unit square in either random locations (Layout =
Random) or gridded locations (Layout = Gridded). The mean for the
spatial process generating the response was set to zero.

@@

Figure \ref{fig:figeff} shows the relative efficiency of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout, where ``IRS-Design'' is the baseline: \mbox{} \begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

@@

We also studied 95\% interval coverage among the approaches. The
design-based 95\% confidence intervals and model-based 95\% prediction
intervals are constructed using the normal distribution. Justification
for the design-based intervals lies in the asymptotic normality of
totals via the Central Limit Theorem, and justification for the
model-based intervals lies in the normality assumption of the errors.
Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four approaches in the random location layout. All four approaches
have somewhat similar interval coverage in all settings, with
GRTS-Design having slightly lower coverage when the response is normal.

In the normal response settings, all approaches have coverage around
95\%. This is expected, as the intervals are also based on the normal
distribution In the lognormal response settings, however, all approaches
have coverage below 95\%. This is also expected, as the intervals are
still based on the normal distribution. In the lognormal response
settings, interval coverage increases both as the sample size increases
and as the strength of spatial dependence increases. This suggests that
the larger the sample size and the stronger the spatial dependence, the
more resistant these intervals are to departures from normality of the
data. These conclusions were similar to those observed when the data
were gridded.
